---
title: "A basic vignette on grids in SOMbrero"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
---

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{A basic vignette on grids in SOMbrero}
%\VignettePackage{SOMbrero}
-->

```{r loading, results='hide', echo=FALSE, warning=FALSE, message=FALSE}
library(SOMbrero)
```

## Definition and purpose of a `myGrid` class object

Objects of class `myGrid` are made to display the SOM grid. 

## Basic functions on a `myGrid` class object

In this section, we will consider only the four basic functions that can be 
applied on a `myGrid` class object:

* `initGrid,`
* `print.myGrid`
* `summary.myGrid`
* `plot.myGrid`

### The `initGrid` function

The `initGrid` function initializes a new `myGrid` object. It has 3 arguments:

* ```dimension```, which is a vector of two integers. The first one is the x
dimension (number of neurons/units on the x axis) and the second one is the y 
dimension. The __default dimensions__ are __x=5__ and  __y=5__,

* ```topo```, which is the chosen topology. The value of this argument must be
one of  ```square``` or `hexagonal`. The__default value__ is __square__.

* ```dist.type```, which is the distance type to compute the distance between
two neurons of the grid. The __default value__ is __euclidean__ and the 
coordinates of the neurons on the grid are chosen so as the minimum Euclidean
distance between two neurons is exactly 1. `hexagonal` topoology is only 
compatible with the Euclidean distance.

The following **R** code initializes a new `myGrid` object of square topology, x 
dimension 5 and y dimension 6, and distance type ```maximum```.

```{r initGird}
first_grid <- initGrid(dimension = c(5,6), topo = "square", dist.type = "maximum")
```

### The `print.myGrid` function

The `myGrid` object `print` function prints the main features of the chosen 
object in the console. The only argument is the object to be printed.

Considering the previously initialized grid, the print command is:

```{r plotFirstGrid}
print(first_grid)
```

### The `summary.myGrid` function

The `myGrid` object `summary` function is quite simple. It only prints the class
of the object and then calls the print function previously described. The only 
argument is the object to be summarized.

```{r summaryMyGrid}
summary(first_grid)
```

### The `plot.myGrid` function

The `myGrid` object `plot` function draws the squared area corresponding to the 
object, in a new graphical window. It has 2 parameters:

* the object to be plotted,

* the color(s) to be used to fill the areas on the graph. The __default colors__ 
correspond to the default `ggplot2` qualitative colors. If only one color is
specified then all neurons are filled with it and, otherwise, the number of 
colors must correspond to the number of neurons on the map. The map is filled 
from the bottom left corner, up the column and then likewise in the next columns
from left to right.

The neuron number is displayed in the neuron, unless the option 
`print.title = FALSE` is used.

```{r plotRainbowGrid}
plot(first_grid)
plot(first_grid, neuron.col = "white")
plot(first_grid, neuron.col = colorRampPalette(c("white", "pink", "red"))(30),
     print.title = FALSE)
```

Hexagonal grids can be displayed similarly:
```{r plotHexa}
second_grid <- initGrid(dimension = c(4, 5), topo = "hexagonal")
plot(second_grid, the.titles = paste0("N", 1:20)) + 
  ggplot2::ggtitle("Hexagonal SOM grid")
```


### Session information

This vignette has been computed with the following environment:
```{r sessionInfo}
sessionInfo()
```