---
title: "Using Self-Organizing Maps with SOMbrero for dissimilarity matrices"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{SOM for dissimilarity data}
%\VignettePackage{SOMbrero}
-->

## Basic package description

`SOMbrero` implements different variants of the Self-Organizing Map algorithm
(also called Kohonen's algorithm). To process a given dataset with the SOM
algorithm, you can use the function `trainSOM()`.

__This documentation only considers the case of dissimilarity matrices.__

```{r loading, results='hide', echo=FALSE, warning=FALSE, message=FALSE}
library("ggplot2")
library("SOMbrero")
```

### Arguments

The ```trainSOM``` function has several arguments, but only the first one is
required. This argument is ```x.data``` which is the dataset used to train the 
SOM. In this documentation, it is passed to the function as a square matrix or
data frame, which entries are dissimilarity measures between pairs of 
observations. The diagonal of this matrix must contain only zeros.

The other arguments are the same as the arguments passed to the ```initSOM```
function (they are parameters defining the algorithm, see ```help(initSOM)```
for further details).

### Outputs

The ```trainSOM``` function returns an object of class ```somRes``` (see 
```help(trainSOM)``` for further details on this class).

### Graphics

The following table indicates which graphics are available for a relational SOM.

| What <br> <br> SOM or SC<br>Type | SOM<br>Energy  <br> <br> <br>| Obs  <br> <br> <br> <br>| Prototypes  <br> <br> <br> <br>| Add  <br> <br> <br> <br>| SuperCluster<br>(no what) <br> <br> <br> | Obs <br> <br> <br> <br> | Prototypes <br> <br> <br> <br> | Add <br> <br> <br> <br> |
|:------------|:--------|:-----|:------------|:-----|:-------------|:-----|:------------|:-----|
| (no type)   | x       |      |             |      |              |      |             |      |
| hitmap      |         | x    |             |      |              | x    |             |      |
| color       |         |      |             | x    |              |      |             |      |
| lines       |         |      | x           | x    |              |      | x           | x    |
| meanline    |         |      |             | x    |              |      |             | x    |
| barplot     |         |      | x           | x    |              |      | x           | x    |
| pie         |         |      |             | x    |              |      |             | x    |
| boxplot     |         |      |             | x    |              |      |             | x    |
| poly.dist   |         |      | x           |      |              |      | x           |      |
| umatrix     |         |      | x           |      |              |      |             |      |
| smooth.dist |         |      | x           |      |              |      |             |      |
| mds         |         |      | x           |      |              |      | x           |      |
| grid.dist   |         |      | x           |      |              |      |             |      |
| words       |         |      |             | x    |              |      |             |      |
| names       |         | x    |             | x    |              | x    |             |      |
| graph       |         |      |             | x    |              |      |             | x    |
| projgraph   |         |      |             | x    |              |      |             | x    |
| grid        |         |      |             |      |              |      | x           |      |
| dendrogram  |         |      |             |      | x            |      |             |      |
| dendro3d    |         |      |             |      | x            |      |             |      |


## Case study: the ```lesmis``` data set

The ```lesmis``` data set is based on the coappearance graph of the characters 
of the novel Les Miserables (Victor Hugo). Each vertex stands for a character 
whose name is given by the vertex label. One edge means that the corresponding 
two characters appear in a common chapter in the book. Each edge also has a 
value indicating the number of coappearances. The co-appearance network has been
extracted by D.E. Knuth (1993).

The ```lesmis``` data contain two objects: the first one,```lesmis```, is an 
```igraph``` object (see [the igraph web page](http://igraph.sourceforge.net/)),
with `r data(lesmis); vcount(lesmis)` nodes and `r ecount(lesmis)` edges.

Further information on this data set is provided with ```help(lesmis)```.

```{r lesmisDescr, fig.width=12, fig.height=12}
data(lesmis)
lesmis
plot(lesmis, vertex.size = 0)
```

The ```dissim.lesmis``` object is a matrix with entries equal to the length of 
the shortest path between two characters (obtained with the function
```shortest.paths``` of package ```igraph```). Note that its row and column
names have been initialized with the characters' names to ease the use of the 
graphical functions of ```SOMbrero```.

### Training the SOM

```{r lesmisTrain, cache=TRUE}
set.seed(622)
mis.som <- trainSOM(x.data=dissim.lesmis, type = "relational", nb.save = 10,
                   init.proto = "random", radius.type = "letremy")
plot(mis.som, what="energy")
```

The dissimilarity matrix ```dissim.lesmis``` is passed to the ```trainSOM``` 
function as input. As the SOM intermediate backups have been registered
(```nb.save = 10```), the energy evolution can be plotted: it stabilized in the
last 100 iterations.

### Resulting clustering

The clustering component provides the classification of each of the 77 
characters. The ```table``` function is a simple way to view data distribution 
on the map.
```{r lesmisClustering}
mis.som$clustering
table(mis.som$clustering)
plot(mis.som)
```

The clustering can be displayed using the ```plot``` function with 
```type = names```.
```{r lesmisPseudoNamesPlot, fig.height=12, fig.width=12, warning=FALSE}
plot(mis.som, what = "obs", type = "names")
```

In this clustering, the main character, Valjean, is in a central position (in 
cluster 8) and some clusters are easily identified as sub-stories around 
Javert. For instance, clusters 10, 15 and 20 are related to the Thénardier 
family, with (for instance), cluster 20 being the cluster of Gavroche and his
two brothers (named children 1 and 2).

The original graph can also be superimposed on the map:
```{r lesmisProjGraph}
plot(mis.som, what = "add", type = "graph", var = lesmis)
```

In the latter plot (which is still messy at this stage of the analysis), nodes
correspond to clusters and are positionned at the cluster location on the map. 
The size of the nodes is proportional to the number of characters classified in
this cluster and edges between nodes have a width proportional to the total 
weight between any two characters from the two linked clusters.

Clusters profile overviews can be plotted either with *e.g.*, lines or barplot, 
that both provide an information similar to that given by `"names"`.
```{r lesmisProtoProfiles}
plot(mis.som, what = "prototypes", type = "lines")  +
  guides(color = guide_legend(keyheight = 0.5, ncol = 2, label.theme = element_text(size = 6))) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
plot(mis.som, what = "prototypes", type = "barplot")  +
  guides(fill = guide_legend(keyheight = 0.5, ncol = 2, label.theme = element_text(size = 6))) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

On these graphics, one variable is represented respectively with a point or a 
slice. It is therefore easy to see which variable affects which cluster.

To see how different the clusters are, some graphics show the distances between 
prototypes. These graphics have exactly the same interpretation as for the other 
data types processed by `SOMbrero`.

* ```"poly.dist"``` represents the distances between neighboring prototypes with
polygons plotted for each cell of the grid. The smaller the distance between a 
polygon's vertex and a cell border, the closer the pair of prototypes. The 
colors encode the number of observations in the neuron;

* ```"umatrix"``` fills the neurons of the grid using colors that represent the 
average distance between the current prototype and its neighbors;

* ```"smooth.dist"``` plots the mean distance between the current prototype and 
its neighbors with a color gradation;

* ```"mds"``` plots the number of the neuron on a map according to a 
Multi-Dimensional Scaling (MDS) projection;

* ```"grid.dist"``` plots a point for each pair of prototypes, with the $x$ 
coordinates representing the distance between the prototypes in the input space,
and $y$ coordinates representing the distance between the corresponding neurons 
on the grid.

```{r lesmisProtoDist}
plot(mis.som, what = "prototypes", type = "poly.dist")
plot(mis.som, what = "prototypes", type = "smooth.dist")
plot(mis.som, what = "prototypes", type = "umatrix")
plot(mis.som, what = "prototypes", type = "mds")
plot(mis.som, what = "prototypes", type = "grid.dist")
```

Here we can see that the prototypes located in the top left and top right 
corners of the map (*e.g.*, clusters 5 and clusters 19-20 and 24-25) are further
from the other neurons than in average.

Finally, with a graphical overview of the clustering
```{r lesmisColorOverview, fig.height=12, fig.width=12}
plot(lesmis, vertex.label.color = rainbow(25)[mis.som$clustering], 
     vertex.size = 0)
legend(x = "left", legend = 1:25, col = rainbow(25), pch = 19)
```

We can see that (for instance) cluster 25 is very relevant to the story: as the 
characters of this cluster appear only in the sub-story of the Bishop 
```Myriel```, he is the only connection for all other characters of cluster 25. 
The same kind of conclusion holds for cluster 20 (with Gavroche), among others. 
Most of the other clusters have a small number of observations: it thus seems 
relevant to compute super clusters.

### Compute super clusters

As the number of clusters is rather large with the SOM algorithm, it is 
possible to perform a hierarchical clustering on top of SOM results. First, let 
us have an overview of the dendrogram:
```{r lesmisSCOverview}
plot(superClass(mis.som))
```

According to the proportion of variance explained by super clusters, 5 groups 
seem to be a good choice (4 groups would have been relevant also. The clustering
with 5 groups creates a group with only one cluster in it).
```{r lesmisSC}
sc.mis <- superClass(mis.som, k = 5)
summary(sc.mis)
table(sc.mis$cluster)
plot(sc.mis)
plot(sc.mis, what = "prototypes", type = "grid")
plot(sc.mis, what = "prototypes", type = "lines")
plot(sc.mis, what = "prototypes", type = "mds")
plot(sc.mis, type = "dendro3d")
```

```{r lesmisSCColorOverview, fig.width=12, fig.height=12}
plot(lesmis, vertex.size = 0, 
     vertex.label.color = rainbow(5)[sc.mis$cluster[mis.som$clustering]])
legend(x = "left", legend = paste("SC", 1:5), col = rainbow(5), pch = 19)
```

* cluster 1 contains ```Myriel``` and the characters involved in his 
sub-story;

* cluster 2 constains ```Valjean``` which has a central position in the graph
visualization, and most of the important character of the novel (including
Javert, Fantine and Cosette);

* cluster 3 countains people almost only connected to ```Fantine``` who links
them to the rest of the novel;

* cluster 4 contains ```Gavroche```, the abandonned child of the
```Thenardier```, and the characters of his sub-story (including Mr Thénardier
and Gavroche's two brothers and his sister, `Eponine`);

* cluster 5 is a bit harder to interpret, with secondary characters related to
`Thenardier` and to the main characters of the novel.

SOMbrero also contains functions to compute a projected graph based on the
super-clusters and to display it:

```{r lesmisSCProjGraph}
projectIGraph(sc.mis, lesmis)
par(mar = rep(0,4))
plot(sc.mis, what = "add", type = "projgraph", variable = lesmis, s.radius = 2)
```

This representation provides a simplified and interpretable display of the graph
where the super clusters are represented by nodes with sizes proportional to the
number of characters classified in them. The nodes are positionned at the center
of gravity of the map clusters included in each super cluster. They are linked
to each other with edges with width proportional to the total number of links 
between two characters of the correspondind super clusters. Here, the central
brown/green node is the one of Valjean and the other main characters (super
cluster 2), which appears to be strongly related to super cluster 4 in blue, 
with Gavroche's neighbors.


### Session information

This vignette has been computed with the following environment:
```{r sessionInfo}
sessionInfo()
```