<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{SOM for contingency tables}
-->

# Self-Organizing Map for contingency tables

#### Laura BENDHAIBA, Madalina OLTEANU, Nathalie VILLA-VIALANEIX

## Basic package description

To be able to run the SOM algorithm, you have to load the package called 
```SOMbrero```. The function used to run it is called ```trainSOM()``` and is 
detailed below.

__This documentation only considers the case of contingency tables.__

```{r loading, results='hide', echo=FALSE, warning=FALSE, message=FALSE}
library(SOMbrero)
```

### Arguments

The ```trainSOM``` function has several arguments, but only the first one is
required. This argument is ```x.data``` which is the dataset used to train the 
SOM. In this documentation, it is passed to the function as a matrix or a data
frame. This set must be a contingency table, i.e., it must contain either 0 or 
positive integers. Column and row names must be supplied.

The other arguments are the same as the arguments passed to the ```initSOM```
function (they are parameters defining the algorithm, see ```help(initSOM)```
for further details).

### Outputs

The ```trainSOM``` function returns an object of class ```somRes``` (see 
```help(trainSOM)``` for further details on this class).

## Case study: the ```presidentielles2002``` data set

The ```presidentielles2002``` data set provides the number of votes at the first
round of the 2002 French presidential election for each of the 16 candidates in
all of the 106 French administrative districts called "departements". Further
details about this data set and the 2002 French presidential election are given
with ```help(presidentielles2002)```.

```{r loadData}
data(presidentielles2002)
apply(presidentielles2002,2,sum)
```
(the two candidates that ran the second round of the election were Jacques
Chirac and the far-right candidate Jean-Marie Le Pen)

### Training the SOM

```{r presiTrain, cache=TRUE}
set.seed(4031719)
korresp.som <- trainSOM(x.data=presidentielles2002, dimension=c(8,8),
                        type="korresp", scaling="chi2", nb.save=10)
korresp.som
```

As the energy is registered during the intermediate backups, we can have a look
at its evolution
```{r energyPresi}
plot(korresp.som, what="energy")
```
which is stabilized during the last 100 iterations.

### Resulting clustering

The clustering component contains the final classification of the dataset. As 
both row and column variables are classified, the length of the resulting 
vector is equal to the sum of the number of rows and the number of columns.

__NB: The clustering component shows first the column variables (here, the 
candidates) and then the row variables (here, the departements).__

The following table indicates which graphics are available for a korresp SOM.

Type        |   Energy  |   Obs |   Prototypes  |   Add   |   Super Cluster
------------|:---------:|:-----:|:-------------:|:-------:|:----------------:
no type     |   x       |       |               |         |
hitmap      |           |   x   |               |         |         x
color       |           |       |       x2      |         |         x2
lines       |           |       |       x2      |         |         x2
barplot     |           |       |       x       |         |
radar       |           |       |       x       |         |
pie         |           |       |               |         |
boxplot     |           |       |               |         |
3d          |           |       |       x2      |         |
poly.dist   |           |       |       x       |         |         x
umatrix     |           |       |       x       |         |
smooth.dist |           |       |       x       |         |
words       |           |       |               |         |
names       |           |   x   |               |         |
graph       |           |       |               |         |
mds         |           |       |       x       |         |         x
grid.dist   |           |       |       x       |         |         
grid        |           |       |               |         |         x
dendrogram  |           |       |               |         |         x
dendro3d    |           |       |               |         |         x

In the column "Prototypes", a plot marked "x2" means that this plot is 
available for both row and column variables. In the "Super Cluster" column, a 
"x2" cell means the plot is available for both data set variables and 
additional variables. 

```{r presiClusters}
korresp.som$clustering
```
The resulting distribution of the clustering on the map can also be visualized
by a hitmap:
```{r presiHitmap}
plot(korresp.som, what="obs", type="hitmap")
```
For a more precise view, ```"names"``` plot is implemented: it prints, 
in each neuron, the names of the variables assigned to it ; in the korresp SOM, 
both row and column variable names are printed.
```{r presiGraphObs, warning=FALSE, fig.width=12, fig.height=12}
plot(korresp.som, what="obs", type="names", scale=c(1.5,0.5))
```

The map is organized as follows: the bottom right side of the map is associated
to the candidate "Taubira" who obtained her better vote scoring in the overseas
departements "Guadeloupe", "Martinique" and "Guyane".

These candidates are opposed to the top left hand side of the map which is 
associated to the far-right candidates "Le Pen" and "Megret" who traditionally
obtain higher voting scores in some South of France departments including 
"Herault", "Vaucluse", "Gard". A little bit further on the right, we can notice 
the candidate "Saint Josse" who was fighting for traditional living style and 
hunting and who obtained his higher scores in the rural departements "Somme", 
"Landes" and "Aude".

In the center of the map are the most traditional candidates "Jospin", "Chirac",
"Hue"... who obtained more uniform scores in the whole of France. It has
to be noted that "Le Pen", who is a far right-candidate, has a rather central 
position also and actually succeeded for the first time to run the second round
of the presidential election.

### Clustering interpretation

Some graphics from the numeric SOM algorithm are still available in the korresp 
case. They are detailed below. As the resulting clustering provides the 
classification for both rows and columns, a new argument ```view``` is used to 
specify which one should be considered. Its possible values are either 
```"r"``` for row variables (the default value) or ```"c"``` for column 
variables.

#### Graphics on prototype values

Three representations are available:

* with lines: either all row or all column variables are plotted (```view``` 
argument is used)
```{r presiProtoL}
# plot the line prototypes (106 French departements)
plot(korresp.som, what="prototypes", type="lines", view="r", print.title=TRUE)
# plot the column prototypes (16 candidates)
plot(korresp.som, what="prototypes", type="lines", view="c", print.title=TRUE)
```
The peaks in neurons 49, 57 and 58 correspond, in the row view, to the 
overseas departement and, in the column view, to the candidate "Taubira". In the
column views, the two peaks clearly identified in the central clusters
correspond to the two "main" tranditional candidates "Jospin" and "Chirac"
(respectively, left and right candidates).

A more precise individual view are given with the graphics "color" and "3d",
here drawn, as an example for the candidate "Le Pen" and for the departement
"Martinique".
* in color: one of the row or column variable (the variable is chosen with the 
argument ```variable```) is represented on the map;
* in 3d, which handling is similar to ```"color"```.
```{r presiProtoC3d, fig.width=12, fig.height=6}
par(mfrow=c(1,2))
plot(korresp.som, what="prototypes", type="color", variable="LE_PEN")
plot(korresp.som, what="prototypes", type="3d", variable="martinique")
```
The first graphic shows that "Le Pen" obtained its best scores in the
departements located in the top left hand side and in the top right hand side of
the map and its lowest scores in the departements located in the bottom right
side of the map (overseas departement).

The second graphic shows that the candidates that obtained the higher scores in
Martinique are located in the bottom right hand side of the map (mainly 
Taubira).

The graphics can also be drawn by giving the variable number and its type,
either "r" or "c" (here, as an example, "Chirac" which is the 5th candidate):
```{r presiProtoNumber, fig.width=12, fig.height=6}
par(mfrow=c(1,2))
plot(korresp.som, what="prototypes", type="color", variable=5, view="c")
plot(korresp.som, what="prototypes", type="3d", variable=5, view="c")
```
Hence "Chirac" has a rather central position on the map which illustrates the
fact that he is actually a "traditional" candidate with almost uniform votes
in the whole country.

#### Graphic on prototype distances

These graphics are exactly the same as in the numerical case:
* ```"poly.dist"``` represents the distances between neighboring prototypes with
polygons plotted for each cell of the grid. The smaller the distance between 
a polygon's vertex and a cell border, the closer the pair of prototypes.
The colors indicates the number of observations in the neuron (white=empty);
* ```"umatrix"``` fills the neurons of the grid using colors that represent
the average distance between the current prototype and its neighbors;
* ```"smooth.dist"``` plots the mean distance between the current prototype and 
its neighbors with a color gradation;
* ```"mds"``` plots the number of the neuron on a map according to a Multi
Dimensional Scaling (MDS) projection;
* ```"grid.dist"``` plots a point for each pair of prototypes, with x 
coordinates representing the distance between the prototypes in the 
input space, and y coordinates representing the distance between the 
corresponding neurons on the grid.
```{r presiGraphProto2}
plot(korresp.som, what="prototypes", type="poly.dist", print.title=TRUE)
plot(korresp.som, what="prototypes", type="umatrix", print.title=TRUE)
plot(korresp.som, what="prototypes", type="smooth.dist", print.title=TRUE)
plot(korresp.som, what="prototypes", type="mds")
plot(korresp.som, what="prototypes", type="grid.dist")
```
Three neurons (49, 57 and 58) have been already picked out in the 
section __Clustering interpretation__ for having prototypes rather different
than the rest of the map. The graphics just above confirm this hypothesis: there
a noticeable peak in prototype distances around these 4 neurons. The MDS
vizualisation also shows that these 4 prototypes are clearly different.

### Analyze the projection quality

```{r presiQuality}
quality(korresp.som)
```
By default, the quality function calculates both quantization and topographic 
errors. It is also possible to specify which one you want to obtain, by using
the argument ```quality.type```.

The topographic error value varies between 0 (good projection quality) and 1 
(poor projection quality). Here, the topographic quality of the mapping is 
quite good with a topographic error equal to 0.14.

The quantization error is an unbounded positive number. The closer from 0 it
is, the better the projection quality is.

### Building super classes from the resulting SOM

In the SOM algorithm, the number of clusters is necessarily equal to the number
of neurons of the grid. This - quite large - number may not suit the original
data for a clustering purpose.

A usual way to address clustering with SOM is to perform a hierarchical
clustering on the prototypes. This clustering is directly available in the
package ```SOMbrero``` using the function ```superClass```. To do so, you can
first have a quick overview to decide on the number of super clusters which 
suits your data.

```{r presiSC1}
plot(superClass(korresp.som))
```
By default, the function plots both a dendrogram and the evolution of the
percentage of explained variance. Here, 3 super clusters seem to be a good
choice. The output of ```superClass``` is a ```somSC``` class object.
Basic functions have been defined for this class:
```{r presiSC2}
my.sc <- superClass(korresp.som, k=3)
summary(my.sc)
plot(my.sc, plot.var=FALSE)
```

Like ```plot.somRes```, the function ```plot.somSC``` has an 
argument ```'type'``` which offers many different plots and can thus be 
combined with most of the graphics produced by ```plot.somSC```:

Case ```"grid"``` fills the grid with colors according to the super clustering 
(and can provide a legend).
Case ```"dendro3d"``` plots a 3d dendrogram.
```{r presiSC3}
plot(my.sc, type="grid", plot.legend=TRUE)
plot(my.sc, type="dendro3d")
```
The 3 different neurons mentionned earlier have been gathered together in the 
super cluster 3.


A couple of plots from ```plot.somRes``` are also available for the super 
clustering. Some identify the super clusters with colors:
```{r presiSC4}
plot(my.sc, type="hitmap", plot.legend=TRUE)
plot(my.sc, type="lines", print.title=TRUE)
plot(my.sc, type="lines", print.title=TRUE, view="c")
plot(my.sc, type="mds", plot.legend=TRUE)
```

And some others identify the super clusters with titles:
```{r presiSC5}
plot(my.sc, type="color", view="r", variable="correze")
plot(my.sc, type="color", view="c", variable="JOSPIN")
plot(my.sc, type="poly.dist")
```

Let us consider the third super cluster. It contains 3 departments and 1 
candidate:
```{r DOM, echo=FALSE}
names(korresp.som$clustering)[which(korresp.som$clustering %in%
                                      which(my.sc$cluster==3))]
```
The departments are the 3 biggest overseas departements. These departements, 
regarding history and culture, are different from metropolitan France thus they 
also have a different but common election behaviour. Particularly, during the 
2002 French presidential election, they strongly promoted Christine Taubira, 
who is actually the candidate assigned to this super cluster, a woman 
originated from one of the overseas departements.
